---
title: "Homework 2"
author: "Zakari L. Billo"
date: "2026-03-15"
output: pdf_document
---

```{r setup, include=FALSE}
library(tidymodels)
library(splines)
library(mgcv)
library(pdp)
library(earth)
library(tidyverse)
library(ggplot2)
library(caret)

knitr::opts_chunk$set(
  fig.width = 8,
  fig.height = 7,
  fig.align = "center",
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

Initialize Data

```{r}
college <- read.csv("data/College.csv",  na = c("NA", ".", "")) |>
  janitor::clean_names() |>
  select(-college) |>
  select(outstate, everything()) |>
  mutate(across(everything(), as.numeric))

summary(college) # check overall structure

# we should cap graduation rate at 100%
college$grad_rate[college$grad_rate > 100] <- 100
college <- na.omit(college)

# matrix of predictors
x <- model.matrix(outstate ~ ., college)[, -1]
# vector of response
y <- college$outstate

# Get the image parameters of the current trellis graph
theme1 <- trellis.par.get()
theme1$plot.symbol$col <- rgb(.2, .4, .2, .5)
theme1$plot.symbol$pch <- 16
theme1$plot.line$col <- rgb(.8, .1, .1, 1)
theme1$plot.line$lwd <- 2
theme1$strip.background$col <- rgb(.0, .2, .6, .2)
# Set the modified parameters as the global style for the trellis graph
trellis.par.set(theme1)
featurePlot(x, y, plot = "scatter", labels = c("", "Y"),
type = c("p"), layout = c(3, 2))
```
Partition the dataset into two parts: training data (80%) and test data (20%).

```{r}
set.seed(123) # make this reproducible 
data_split <- initial_split(college, prop = 0.8)
train_data <- training(data_split)
test_data  <- testing(data_split)
```

(a) Fit smoothing spline models to predict out-of-state tuition (Outstate) using Top10perc as the only predictor, across a range of degrees of freedom. Plot the fitted curve for each degree of freedom. Describe the patterns you observe as the degrees of freedom change. Choose an appropriate degree of freedom for the model and plot this optimal fit. State the
criteria you used to select the degree of freedom.

```{r}
set.seed(123)
pred.grid <- seq(from = -20, to = 120, by = 1)

# predictor and response matrices
x <- train_data$top10perc
y <- train_data$outstate

# smoothing splines with different df
fit3  <- smooth.spline(x, y, df = 3)
fit5  <- smooth.spline(x, y, df = 5)
fit7  <- smooth.spline(x, y, df = 7)
fit10 <- smooth.spline(x, y, df = 10)
fit_opt <- smooth.spline(x, y)
fit_opt$df # optimal df is 4.394647

plot(x, y,
     col = "grey",
     pch = 16,
     xlab = "top10perc (%)",
     ylab = "outstate ($)")

lines(predict(fit3), col = "black", lwd = 2)
lines(predict(fit5), col = "green", lwd = 2)
lines(predict(fit7), col = "orange", lwd = 2)
lines(predict(fit10), col = "red", lwd = 2)
lines(predict(fit_opt), col = "purple", lwd = 2)

legend("topleft",
       legend = c("df=3","df=5","df=7","df=10", "df=4.39"),
       col = c("black","green","orange","red", "purple"),
       lwd = 2)
```

As the df increases from 3 to 10, the natural spline gets more flexible, resulting in overfitting. As determined by the optimal smoothing spline, the appropriate degree of freedom was 4.39 (generalized cross-validation is used to select the degree of freedom or the trace of the smoother matrix).

(b) Train a multivariate adaptive regression spline (MARS) model to predict the response variable. Report the regression function. Present the partial dependence plot of an arbitrary predictor in your model. Report the test error.

```{r}
set.seed(123)

# matrix of predictors
x <- model.matrix(outstate ~ ., train_data)[, -1]
# vector of response
y <- train_data$outstate

ctrl1 <- trainControl(method = "cv", number = 10) # 10-fold cv

mars_grid <- expand.grid(degree = 1:3, nprune = 2:15)

mars.fit <- train(x, y,
  method = "earth",
  tuneGrid = mars_grid,
  trControl = ctrl1)
  ggplot(mars.fit)
```

Let's try to get a sense of the selected hyperparameters

```{r}
mars.fit$bestTune # nprune = 14, df = 1
```
```{r}
coef(mars.fit$finalModel)
print(mars.fit$finalModel)
```
The regression function is $$\begin{aligned}
\hat{y} &= 12974.7011817 \\
&\quad - 0.7523615 \cdot h(\text{expend} - 15954) \\
&\quad - 25.1664857 \cdot h(85 - \text{grad\_rate}) \\
&\quad - 1.3090278 \cdot h(4440 - \text{room\_board}) \\
&\quad - 1.0904120 \cdot h(\text{f\_undergrad} - 1365) \\
&\quad - 2.2469701 \cdot h(1365 - \text{f\_undergrad}) \\
&\quad - 83.5697756 \cdot h(22 - \text{perc\_alumni}) \\
&\quad + 0.3613673 \cdot h(\text{apps} - 1377) \\
&\quad - 2.9673230 \cdot h(\text{enroll} - 213) \\
&\quad + 13.5624272 \cdot h(213 - \text{enroll}) \\
&\quad + 0.7449182 \cdot h(\text{expend} - 6951) \\
&\quad + 1.1719119 \cdot h(\text{f\_undergrad} - 2993) \\
&\quad + 0.4147319 \cdot h(\text{accept} - 1540) \\
&\quad - 1.8564357 \cdot h(1540 - \text{accept})
\end{aligned}$$
```{r}
p1 <- pdp::partial(mars.fit, pred.var = "enroll", grid.resolution = 10) |> 
  pdp::plotPartial()

p1
```
This is the partial dependence plot on the estimated response variable with the predictor `enroll`. Increasing enrollment decreases the predicted out of state tuition. 
```{r}
mars_preds <- as.numeric(predict(mars.fit, newdata = test_data))
test_mse <- mean((test_data$outstate - mars_preds)^2, na.rm = TRUE) 
print(paste("MARS Test MSE:", round(test_mse, 2)))
print(mars.fit$finalModel)
```
The MARS test MSE on the response variable `outstate` was 2,589,498.05. MARS selected 14 of 30 terms, and 8 of 16 predictors (nprune=14).

(c) Construct a generalized additive model (GAM) to predict the response variable. For the nonlinear terms included in your model, generate plots to visualize these relationships and discuss your observations. Report the test error.

```{r}
set.seed(123)

gam.mod <- gam(
  outstate ~ s(apps) + s(accept) + s(enroll) + s(top10perc) + s(top25perc) +
    s(f_undergrad) + s(p_undergrad) + s(room_board) + s(books) +
    s(personal) + s(ph_d) + s(terminal) + s(s_f_ratio) +
    s(perc_alumni) + s(expend) + s(grad_rate),
  data = train_data
)

# Model summary
gam_sum <- summary(gam.mod)

# Find smooth terms (edf > 1)
nonlinear_idx <- which(gam_sum$s.table[, "edf"] > 1.001)

# See the smooth terms 
rownames(gam_sum$s.table)[nonlinear_idx]
```
The smooth or nonlinear terms are `apps`, `accept`, `f_undergrad`, `room_board`, `books`, `personal`, `ph_d`, `terminal`, `s_f_ratio`, `perc_alumni`, and `expend`. 
```{r}
# Plot only nonlinear / smooth terms
par(mfrow = c(3, 4))
for (i in nonlinear_idx) {
  plot(gam.mod,
       select = i,
       shade = TRUE,
       seWithMean = TRUE)
}
```
The terms `apps`, `accept`, `f_undergrad`, `room_board`, `ph_d`, `s_f_ratio`, `perc_alumni`, and `expend` have the most nonlinear relationships with the response variable `outstate`. 

```{r}
set.seed(123)

# MSE
gam.pred = predict(gam.mod, newdata = test_data)
y2 = test_data$outstate
mean((y2 - gam.pred)^2)
```
The test MSE for the GAM model is 2,298,158.

(d) In this dataset, would you favor a MARS model over a linear model for predicting out-ofstate tuition? Provide a clear justification for your choice.

```{r}
set.seed(123)

lm.fit <- train(
  outstate ~ .,
  train_data,
  method = "lm",
  trControl = ctrl1
)

# compare cross-validated results
resamp <- resamples(list(
  mars = mars.fit,
  lm   = lm.fit
))

summary(resamp)

# boxplot of RMSE
bwplot(resamp, metric = "RMSE")
```
```{r}
# evaluate the test MSE for linear model
lm.pred <- predict(lm.fit, newdata = test_data)
lm_mse <- mean((test_data$outstate - lm.pred)^2)
print(paste("Linear Model Test MSE:", round(lm_mse, 2)))
```
In this data I would favor the multivariate adaptive regression spline (MARS) model. The linear model has a test MSE of 3,318,588.69 compared to the MARS's test MSE of 2,589,498.05. The fact that the linear model performs worse on the testing (and training data makes sense); the MARS model can account for more nonlinearity in the relationship between the response variable `outstate` and the predictors. 

(e) In practical applications, would you consider a MARS model to be superior to a linear model, or the other way around? Please justify your answer with appropriate reasoning.

If it is unjustified to believe that there is nonlinearity in the relationship between the response variable and the predictors, or if the linear model provides similar test error, a linear model is probably the better choice on the basis of interpretability. The multivariate adaptive regression spline (MARS) model is more helpful when the the data show curvature or threshold effects, because MARS models build a piecewise linear model using hinge basis functions and will automatically capture nonlinear predictor-response relationships.